---
title: "FAOSTAT_Cropgrids II "
output: html_notebook
---

# Backcasting CROPGRIDS from 2020 to 2000 II

IDEA Assume distribution remains more or less the same adjust sum of crops with FAOSTAT data

set paths to dir, read packages etc

```{r}
library(dplyr)
library(terra)
library(tidyr)
library(readxl)
library(countrycode)
library(readr)
data_dir <- "data/"
```

## Clean FAOSTAT data

# downloaded for all years

```{r}
FAO_csv <- read.csv(paste0(data_dir,
                    "05_crop_types/CG/FAOSTAT/Production_Crops_Livestock_E_All_Data/Production_Crops_Livestock_E_All_Data.csv"
))

# clean this a bit 
FAO_cleaned <- FAO_csv |>
  filter(Element == "Area harvested") |>
  filter(Area.Code < 5000 ) |> # eliminate all the Continents / world region figures

  select(
    # Keep metadata columns (adjust as needed)
    `Area.Code`, `Area.Code..M49.`, Area, `Item.Code`, `Item.Code..CPC.`, Item,
    Element, Unit,
    # Select year columns from 2000 onwards excluding F and N
    matches("^Y(20(0[0-9]|1[0-9]|2[0-9]|30))$")
    )|>  
  mutate(
    area_code_M49 = parse_number(`Area.Code..M49.`),
    iso3 = countrycode(area_code_M49, "un", "iso3c"),
    iso3 = case_when(
      area_code_M49 == 158 ~ "TWN",
      TRUE ~ iso3
  ))


table_crops_matching <- read_xlsx(paste0(data_dir, "01_raw/CROPGRID/CODES/CROPS_matching_v17.xlsx"))

table_crops_matching_unique <- table_crops_matching |>
  distinct(`FAO NAME`, .keep_all = TRUE) |>
  select(`FAO NAME`, `REMAP MRF NAME`)

# merge both tables to have the final crop name in the dataset.

FAO_cleaned <- FAO_cleaned |>
  left_join(table_crops_matching_unique, by = join_by(Item == `FAO NAME`))

```

--\> I just assigned the existing FAOSTAT Items to the classes from CROPGRIDS. Next step:

Add the "new" Item code to the table and bring it to a long format

```{r}
crops_match2 <- readr::read_csv2(paste0(data_dir,
                                      "/05_crop_types/CG/FAOSTAT/FAOSTAT_codes/CROPS_matching_allyears.csv"))
# transfer FAO_cleaArea.Code# transfer FAO_cleaned to a long format
# ts stands for timeseries 
FAO_cleaned2 <- FAO_cleaned |>
  left_join(crops_match2, by = join_by(Item == FAO))

FAO_long_ts <- FAO_cleaned2 |>
  filter(Area.Code < 5000 ) |> # eliminate all the Continents / world region figures
  select(Area, area_code_M49,iso3, Item,CROPGRIDS, Y2000, Y2001, Y2002,Y2003, Y2004, Y2005, Y2006, 
         Y2007, Y2008, Y2009, Y2010, Y2011, Y2012, Y2013, Y2014, Y2015,
         Y2016, Y2017, Y2018, Y2019, Y2020, Y2021, Y2022, Y2023,Y2024)|>
  pivot_longer(cols = starts_with("Y"), 
               names_to = "Year", 
               values_to = "area_ha", 
               values_drop_na = T) |>
    mutate(Year = as.factor(gsub("Y", "", Year)))|>
  filter(CROPGRIDS != "No") |>
  filter(Area != "China")

head(FAO_long_ts)

```

Looks good. Safe it as csv

```{r}
write.csv(FAO_long_ts, paste0(data_dir, "/05_crop_types/CG/FAOSTAT/FAOSTAT_codes/FAOSTAT_cleaned_2000-2024.csv"), 
          row.names = F)
```

# Calculate change for each year to baseline

Baseline is 2020

## Check baseline data from CG Codes vs downloaded FAOSTAT data

```{r}
FAO_20_org <- read_xlsx("data/01_raw/CROPGRID/CODES/FAOSTAT_ALL_2020_edited.xlsx")
# combine with cropgrids 
FAO_20_org <- FAO_20_org |>
  left_join(table_crops_matching_unique, join_by(Item == `FAO NAME`))

sum(FAO_20_org$Value, na.rm=T)

FAO_base <- FAO_long_ts |>
  filter(Year == 2020)

sum(FAO_base$area_ha)

# check where the difference is? 

FAO_b_comb <- FAO_base |>
  group_by(Area, CROPGRIDS) |>
  summarize(area_ha = sum(area_ha, na.rm=T))|>
  left_join(FAO_20_org, by = join_by(Area == Area, CROPGRIDS == `REMAP MRF NAME`))|>
  select(Area,CROPGRIDS, Value, area_ha)

FAO_b_comb$diff <- FAO_b_comb$area_ha - FAO_b_comb$Value

# summarize by croptype 
FAOct_diff <- FAO_b_comb |>
  group_by(CROPGRIDS) |>
  summarize(area_ha = sum(area_ha, na.rm=T), 
            Value = sum(Value, na.rm=T), 
            diff = sum(diff, na.rm=T))
```

--\> in some cases, the values do not align, this is when Tian et al. chose national datasets to correct FAOSTAT statistics --\> here we will just take FAOSTAT data for the sake of simplification.

## Calculate the change per crop and country to baseline

```{r}
FAO_long_change <- FAO_long_ts |>
  left_join(FAO_base, by = join_by(Item == Item, CROPGRIDS == CROPGRIDS, iso3 ==iso3,Area == Area,area_code_M49 ==area_code_M49))|>
  rename(base_area = area_ha.y, 
         area_y = area_ha.x, 
          year =Year.x)|>
  select(!Year.y) |>
  mutate(
    bbase_area = coalesce(base_area, area_y),              # replace NAs
    bbase_area = if_else(bbase_area == 0, 1e-6, bbase_area), # avoid division by zero
    change_prop = area_y / bbase_area,
    change_absolute = area_y - bbase_area
  )

```

### Remove country-crop combination with 0 values from 2020 
Division by 0 --> some crops in some countrys report 0 harvest in 2020 but do report in previous years 
Problem now: Cannot spatially backcast them using CROPGRIDS --> remove those crops from the countries 

```{r}
FAO_long_change <- FAO_long_change |>
  filter(base_area!=0)
```

## Create change-raster per crop-type

Now start with the tricky part: Backcasting this to the rasterdataset

first: load country-raster

```{r}
country_rast <- rast("data/04_bia_inputs/LUH2/country_raster.tif")
shp_country <- vect("H:/02_Projekte/allgemein_biodiversity_impact/02_data/country_shp/ne_110m_admin_0_countries.shp")
countries_df <- data.frame(shp_country)
countries_df$country_code<- c(1:nrow(countries_df))
shp_country$country_code<- c(1:nrow(shp_country))

# Use existing ISO3 field from Natural Earth
countries_df$country_id <- as.numeric(factor(countries_df$ISO_A3))

country_lookup <- countries_df[, c("country_id", "ISO_A3", "NAME")]


# merge this table to the change table 
country_name_map <- tibble::tibble(
  FAO = c("Bolivia (Plurinational State of)", "Viet Nam", "Côte d'Ivoire", "Iran (Islamic Republic of)", 
          "Lao People's Democratic Republic", "Republic of Korea", "Democratic People's Republic of Korea", 
          "Democratic Republic of the Congo", "Republic of Moldova", "Russian Federation", 
          "Syrian Arab Republic", "United Republic of Tanzania", 
          "United Kingdom of Great Britain and Northern Ireland", 
          "Brunei Darussalam", "China, mainland", "China, Taiwan Province of", 
          "China, Hong Kong SAR", "China, Macao SAR", "Micronesia (Federated States of)", 
          "Venezuela (Bolivarian Republic of)", "Eswatini", "The Bahamas", "North Macedonia"),
  shp = c("Bolivia", "Vietnam", "Ivory Coast", "Iran", 
          "Laos", "South Korea", "North Korea", 
          "Democratic Republic of the Congo", "Moldova", "Russia", 
          "Syria", "Tanzania", 
          "United Kingdom", 
          "Brunei", "China", "Taiwan", 
          "Hong Kong", "Macau", "Micronesia", 
          "Venezuela", "eSwatini", "Bahamas", "North Macedonia")
)

FAO_long_change <- FAO_long_change |>
  left_join(country_name_map, by = c("Area" = "FAO")) |>
  mutate(country = coalesce(shp, Area)) |>
 select(-shp) |>
left_join(
  countries_df |> select(ISO_A3, country_code),
  by = c("iso3" = "ISO_A3")
)



# get area per year 
print(FAO_long_change |>
      group_by(year)|>
      summarize(area_ha = sum(area_y, na.rm=T)))
```
### add information whether treecrop or not 
```{r}
treecrops <- c("abaca", "agave", "almond", "apple", "apricot", "areca", "avocado", 
               "banana", "carob", "cashew", "cashewapple", "cherry", "chestnut", 
               "citrusnes", "coconut", "cocoa", "coffee", "date", "fig", "fruitnes", 
               "grape", "grapefruitetc", "hazelnut", "karite", "kiwi", "kolanut", 
               "lemonlime", "mango", "oilpalm","olive", "orange", "papaya", "peachetc", "pear", 
               "persimmon", "pimento", "pineapple", "pistachio", "plantain", "plum", 
               "quince", "rasberry", "rubber", "sourcherry", "stonefruitnes", 
               "tea", "tung", "vanilla", "walnut")

FAO_long_change$type <- "crop"
FAO_long_change$type[FAO_long_change$CROPGRIDS %in% treecrops] <- "treecrop" 
```

```{r}
write.csv(FAO_long_change, paste0(data_dir, "/05_crop_types/CG/FAOSTAT/FAOSTAT_codes/FAOSTAT_change_2000-2024.csv"), 
          row.names = F)
```


### save change dataset, further proceed in python with script I_01_02_2 


```{r}
aniseetc <- FAO_long_change |>
  filter(CROPGRIDS == "aniseetc"&
           year == 2019)

sum(aniseetc$area_y, na.rm=T)

# Prepare raster for python processing
sample_CG <-  rast("data/01_raw/CROPGRID/CROPGRIDSv1.08_NC_maps/CROPGRIDSv1.08_yautia.nc")
# prepare a country raster based on FABIO (from Shuntian)
names(shp_country)
country_shp_rast <- rasterize(shp_country, sample_CG,field = "country_code",touches = T)
plot(country_shp_rast)
writeRaster(country_shp_rast, "data/05_crop_types/CG/country_raster_CG.tif", overwrite=T)

# write raster to take this as a blank for python 
blanko_r <- rast(country_shp_rast, vals = 0)
writeRaster(blanko_r,"data/05_crop_types/CG/blanko_raster.tif")

```

# now proceed in Python!!! 


